---
title: "Wages and Education of Young Males"
author: "Richard Yoder, Johnny Sandy, Kyle Sanavullah"
date: "4/21/2022"
output: pdf_document
---
```{r, message=FALSE}
setwd("C:/Users/Owner/Documents/R/BIOL 312 Biostatistics for Bioinformatics/Final Project")

cat("\014") # Clears Console
rm(list=ls()) # Clears Environment
dev.off() # Clears Plots
graphics.off() # Clears Plots if first one didn't work.
```

#Libraries
```{r, message=FALSE}
library(tidyverse)
library(dplyr)
library(modelr)
```

#Data
```{r, message=FALSE}
rm(list=ls()) # Clears Environment
Males <- read_csv("Males.csv")
```

#Cleaning
##Data headings
```{r}
Males <- Males %>% 
  rename(Person_ID = "nr",
         experience = "exper",
         ethnicity = "ethn",
         married = "maried",
         union_membership = "union")
```

#Setting Data to tibble
```{r}
#Males<- tidy(Males)
```

#Exploratory Data Analysis
Is there a relationship with Marriage and Wage?
```{r}
Males %>% 
  ggplot(aes(x = married,
             y = wage))+
  geom_boxplot()
```
Medians and range do appear significantly different.

Let's make data synchronic:
```{r}
temp <- Males %>%
  replace_na() %>%
  filter(year == "1980")

mean(temp$wage)

Males %>% 
  replace_na() %>% 
  filter(year == "1980") %>% 
  select(wage) %>% 
  summary()
```

```{r}
Males %>% 
  group_by(year) %>% 
  summarize(Mean = mean(wage)) %>% 
  ggplot(aes(x = year,
             y = Mean))+
  geom_line()
```

#Questions to explore about dataset

##Question 1: How does each males' wage change from beginning of study (1980) and end of study (1987)

###Check assumption:

1) Random sampling - yes assummed to be random
2) Independent - no, that is why we are doing paired t test
3) Equal variances - Variances do not appear to be equal.
4) Normality - From the qq plot, it appears that year 1980 is non normal.

```{r}
# extrapolate the data from each year
males1980 <- Males %>%
  filter(year == 1980)

males1987 <- Males %>%
  filter(year == 1987)

# plot qq plots
males1980 %>%
  ggplot(aes(sample = wage)) +
  geom_qq() +
  geom_qq_line()



var(males1980$wage)
var(males1987$wage)

wilcox.test(males1980$wage, males1987$wage, paired = TRUE)
```
The p value was 2.2e-16 which is very small compared to alpha = 0.05. Therefore we reject the null hypothesis. The wage difference from the beginning (1980) and the end (1987) is significant.
<<<<<<< Updated upstream
=======

##Question2: If a male is married, is his wage higher than those that are not married?


For sake of independence of the tests, the year 1987 will be used.

###Check assumption:

```{r}
males1987 <- Males %>%
  filter(year == 1987)

males1987 %>%
  ggplot(aes(wage, color = married, fill = married)) + 
  geom_density(alpha = 0.05)

males1987 %>%
  group_by(married) %>%
  summarise(variance = var(wage))

males1987 %>%
  ggplot(aes(sample = wage)) +
  geom_qq() +
  geom_qq_line() + 
  facet_wrap(~married)
```

1) Random sampling - yes assummed to be random
2) Independent - Yes, from years 1987
3) Equal variances - Variances do not appear to be equal.
4) Normality - qq plots show that the data is relatively equal

Test to be used: t test with unequal variance

Null Hypothesis: Mean of married men equal mean of unmarried men
Alternative Hypothesis: Mean of married men do not equal mean of unmarried men

alpha = 0.05
```{r}
t.test(wage ~ married, data = males1987, var.equal = TRUE)
```
The p value waS 0.001033 which is less than alpha therefore this test was significant. Therefore if a man is married, then their wages are higher than those that are not married.

##Question 2: Does different residence equate to higher wages?

Null hypothesis: There is no difference in the means between the residences
Alternative hypothesis: There is at least one of the means that differ between the residences

alpha 0.05

ANOVA assumptions
Random Sampling - met
Residuals are normally distributed - met with qq plot
the variance of the groups have to be equivalent. - The variances are equivalent

Use Welch's one way anova test.
```{r}
noNAResidence <- males1987 %>%
  drop_na(residence)

residenceModel <- lm(wage ~ residence, data = noNAResidence)
plot(residenceModel)

anova(residenceModel)
```

Conclusion: The p value was 0.002079 which is less than alpha = 0.05. Therefore we can reject the null hypothesis. At least one of the residence's mean differes.

Lets figure out which one is different.
```{r}
TukeyHSD(aov(residenceModel))
plot(TukeyHSD(aov(residenceModel)))
```

northern central is higher than north east
south is higher than north east.

##Question 3: Do white individuals have a higher wage than those of any other ethnicities?

```{r}
whiteMales <- males1987 %>%
  filter(ethnicity == "white")
```

## Creating a linear model based on if years experienced to wage

```{r}
wageModel <- lm(wage ~ experience, data = males1987)
```

Assumptions of linear regression

random sampling - met
residuals are independent - true
residuals are normally distributed - qq plot shows that the data relatively normal.
residuals show constant variance - Variances are not equal because the fitted vs residual graph has a funnel. 
x and y have a linear relationship - True, there is a linear relationship.

```{r}
# checking the assumptions
plot(wageModel)

# add residuals to the data set
males1987 <- males1987 %>%
  add_residuals(wageModel, var = "resid_mod1") %>% 
  add_predictions(wageModel, var = "pred_mod1")

males1987 %>%
  ggplot(aes(resid_mod1, wage)) +
  geom_point()
```


```{r}
summary(wageModel)
```
TODO: Need help finishing this, as the variances are not constant.
>>>>>>> Stashed changes
